---
title: "fransoza_preprocessing"
output: html_document
date: "2024-09-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Install required packages (on first use):
```{r}
install.packages('readxl')
install.packages("vegan")
```

#Load required libraries:
```{r}
library(tidyverse)
library(readxl)
library(tidyr)
library(chemometrics)
library(vegan)
library(tableone)
```

# Load the data:
```{r}
data <- load("franzosa_IBD") #main dataset
genus_names <- read_excel('genus_names.xlsx') #genus names dataset
```

##MICROBE DATA:

#Data manipulation:
```{r}
#make row names sample names and remove that row:
rownames(genera) <- genera[,1]
genera <- genera[,-1]
#make column names the genus names:
colnames(genera) <- genus_names$Genus

#create new column 'Disease Group' which is just IBD or Control:
for (val in 1:220){
  if (metadata$Study.Group[val] == "Control"){
    metadata$Disease.Group[val] <-  "Control"
  }
  else {
    metadata$Disease.Group[val] <- "IBD"
  }
}
```

#Data Sparsity:
Remove any features where there are greater than 80% sparsity:
```{r}
sparsity <- c(rep(0,11721)) #empty vectors
#calculate the percentage of zeros for each cluster type for all the samples:
for (i in 1:ncol(genera)) {
    sparsity[i] <- (sum(genera[, i] == 0)/220)*100
}
remove <- which(sparsity > 80)
genera_alt <- genera[-c(remove)]
```
Leaves only now 3960 genera compared to the original 11720. 

#Remove Unknown Family/Genus:
```{r}
#make reduced data a dataframe with numbers only:
genera_alt <- as.data.frame(genera_alt)
genera_alt[] <- lapply(genera_alt, as.numeric)

#genus name of those to keep are listed in column 4 of genus_names:
genera_known <- as.vector(genus_names[1:3268,4])
genera_known <- unlist(genera_known)
genera_alt <- genera_alt[,c(genera_known)]
```
Now 3268 genus are left compared to 3960 previously. Most with genus known, some with family known.

#Find outliers and produce a PCoA:

Scale and Transform the data:
```{r}
#CLR transform:
tran_genera <- clr(genera_alt+0.00001)

#Min-Max Normalisation:
# Define Min-Max normalization function:
min_max_norm <- function(x) {
    (x - min(x)) / (max(x) - min(x))
}

# Apply Min-Max normalization to the dataset:
genera_norm <- as.data.frame(lapply((tran_genera), min_max_norm))
#rename row names as sample names:
rownames(genera_norm) <- metadata[,2]
```

Calculate Brays-Curtis Distance metric:
```{r}
dd <- vegdist(genera_norm, method = "bray")
pcoa_data <- cmdscale(dd, eig = TRUE)
pcoa_df <- data.frame(PC1 = c(pcoa_data$points[,1]),
                      PC2 = c(pcoa_data$points[,2]),
                      DiseaseGroup = metadata$Disease.Group)
```

Plot:
```{r}
ve <- pcoa_data$eig/sum(pcoa_data$eig)
#Plot points:
ggplot(pcoa_df, aes(x = PC1, y = PC2, color = DiseaseGroup, fill = DiseaseGroup)) +
  #change theme:
  theme_classic() +
  #add confidence intervals:
  stat_ellipse(type = "t",level = 0.90, geom= "polygon", alpha = 0.2, show.legend = FALSE)+
  geom_point() +
  #Organize and format figure:
  xlab(paste0("PCo 1 (",round(ve[1]*100,1),"%)")) +
  ylab(paste0("PCo 2 (",round(ve[2]*100,1),"%)")) +
  theme(legend.justification=c(0,0), legend.position=c(0.02,0.02),
        legend.text = element_text(size = 12),
        legend.background = element_rect(fill="gray90", size=1, linetype="dotted"),
        legend.title = element_text(size=13),
        axis.title=element_text(size=14))+
    scale_color_discrete(name="Study Group", labels=c("Control (n = 56)", 
                                                      "IBD (n = 164)"))+
  scale_fill_discrete(name="Study Group", labels=c("Control (n = 56)", "IBD (n = 164)"))

#Save figure:
#ggsave(path = "figs", filename = "pcoa_fran.png")
```

Find outliers and remove them from dataset:
```{r}
#identify outliers:
outliers <- which(pcoa_df$PC2 > 0.2)
#remove:
genera_diff <- genera_alt[-c(outliers),]
metadata_diff <- metadata[-c(outliers),]
```

PERMANOVA comparison between groups:
```{r}
perms <- 1000
alt <- which((metadata$Disease.Group == "IBD")| (metadata$Disease.Group == "Control"))
meta_alt <- metadata[c(alt),]
con_cd <- genera_norm[c(alt),]
ibd_con <- adonis2(
con_cd ~ meta_alt$Disease.Group,
method = "bray", 
permutations = perms
)
pairwise_p <- ibd_con$`Pr(>F)`[1]
p.adjust(pairwise_p, method = "BH")
```


#Reduce features using Mann-Whitney- u:
```{r}
#Obtain indices for each group:
ibd_index <- which(metadata_diff$Disease.Group == "IBD")
control_index <- which(metadata_diff$Disease.Group == "Control")

#complete test for each feature:
man_p <- c(rep(0,3268))
for (feat in 1:3268){
  x <- genera_diff[c(ibd_index), feat]
  y <- genera_diff[c(control_index), feat]
  man_wit <- wilcox.test(x, y)
  man_p[feat] <- man_wit$p.value
}

#adjust for multiple testing:
man_p_adj <- p.adjust(man_p, method = 'hochberg', n = length(man_p))
sig_feat <- which(man_p_adj < 0.05)
#Select only significant features:
genera_diff <- genera_diff[c(sig_feat)]
```
Reduces number of genera from 3268 to 430.

## Extract data:
```{r}
#Genera:
write.csv(genera_diff, "fran_genera.csv", row.names = TRUE)
#Metadata:
write.csv(metadata_diff[, c(2,4,5,9,10,11,12,13,14)],"fran_metadata.csv", row.names = TRUE) # only sections of interest
```

#METABOLITE DATA:

#Data manipulation:
```{r}
#make row names sample names and remove that row:
rownames(mtb) <- mtb[,1]
mtb <- mtb[,-1]
```

#Data sparsity:
Remove any metabolites where there is greater than 80% sparsity:
```{r}
sparsity_mtb <- c(rep(0,8849)) #empty vectors
#calculate the percentage of zeros for each cluster type for all the samples.
for (i in 1:ncol(mtb)) {
    sparsity_mtb[i] <- (sum(mtb[, i] == 0)/220)*100
}
remove_mtb <- which(sparsity_mtb > 80)
mtb_alt <- mtb[-c(remove_mtb)]
```
Leaves now 8835 metabolites compared to the original 8848.

#Remove Unknown Metabolite Compound/Class:
```{r}
#make reduced data a dataframe with numbers only:
mtb_alt <- as.data.frame(mtb_alt)
mtb_alt[] <- lapply(mtb_alt, as.numeric)

#make row names the sample name and remove this row:
rownames(mtb.map) <- mtb.map[,1]
mtb.map <- mtb.map[,-1]

#remove the metabolites with no compund name or class:
remove <- which((is.na(mtb.map$HMDB))&(is.na(mtb.map$KEGG))&is.na(mtb.map$Compound.Name)&is.na(mtb.map$Putative.Chemical.Class))
mtb_alt <- mtb_alt[,c(-remove)]
```
Now 3935 metabolites are left compared to 8835 previously. 

#Find outliers:
Scale and Transform the data:
```{r}
#CLR transform:
tran_mtb <- clr(mtb_alt+ 1)

#z-normalisation:
mtb_norm <- as.data.frame(scale(tran_mtb))
rownames(mtb_norm) <- metadata[,2]
```

Remove same outlier samples as genera data:
```{r}
met_diff <- mtb_alt[-c(126,182),]
metadata_diff <- metadata[-c(126,182),]
```

#Reduce features using Mann-Whitney- u:
```{r}
#obtain indices for each group:
ibd_index <- which(metadata_diff$Disease.Group == "IBD")
control_index <- which(metadata_diff$Disease.Group == "Control")

#complete test for each feature:
man_p <- c(rep(0,3935))
for (feat in 1:3935){
  x <- met_diff[c(ibd_index), feat]
  y <- met_diff[c(control_index), feat]
  man_wit <- wilcox.test(x, y)
  man_p[feat] <- man_wit$p.value
}

#adjust for multiple testing:
man_p_adj <- p.adjust(man_p, method = 'hochberg', n = length(man_p))
sig_feat <- which(man_p_adj < 0.05)
#select significant features only:
met_diff <- met_diff[c(sig_feat)]
```
Reduces from 3935 metabolites to 1343.

#Extract data:
```{r}
write.csv(met_diff, "fran_met.csv", row.names = TRUE)
```

#Analysing Baseline Charactersitics:
```{r}
#Compare Categorical variables:
CreateCatTable(vars = c("mesalamine", "antibiotic", "immunosuppressant", "steroids"), strata = c("Disease.Group"), data = metadata_diff, test = T, addOverall = T)

#Compare Continuous variables (age):
t_Test <- t.test(Age ~ Disease.Group,
           data = metadata_diff)
tapply(metadata_diff$Age, metadata_diff$Disease.Group, summary)
t_Test$p.value

#Summarize age statistics:
mean(metadata_imp$Age, na.rm=TRUE)
sd(metadata_imp$Age, na.rm=TRUE)
tapply(metadata_imp$Age, metadata_imp$Study.Group, sd, na.rm=TRUE)
```
