---
title: "val_hmp2_preprocessing"
output: html_document
date: "2024-09-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Install required packages (on first use):
```{r}
install.packages('readxl')
```

#Load required libraries:
```{r}
library(tidyverse)
library(readxl)
library(tableone)
library(tidyr)
library(car)
```

# Load the data:
```{r}
data <- load("IBD_hmp2") #validation dataset
genus_names <- read_excel('hmp_genus_names.xlsx') #genus names for validation data
```

#Remove paediatric patients:
```{r}
#remove patients with age under 18 at consent:
paed <- which(metadata$consent_age < 18)
metadata_alt <- metadata[-c(paed),]
#remove patients with age under 18 at diagnosis:
paed_diag <- which(metadata_alt$`Age at diagnosis` < 18)
metadata_alt <- metadata_alt[-c(paed_diag),]
#remove unnecessary columns:
metadata_alt <- metadata_alt[,-c(6,7,8,10,20,21,22,23,25,26)]
```

#Get Baseline sample of each patient only:
```{r}
#find the number of patients:
subjects <- unique(metadata_alt$Subject)
#create dataframe to store unique samples:
metadata_uni <- data.frame()
#for each patient extract a subset of all their samples
for (sub in 1:length(subjects)){
  subby <- subset(metadata_alt, Subject == subjects[sub])
  min_week <- min(subby$week_num) #select the sample from the earliest week
  uni_sub <- subset(subby, week_num == min_week)
  metadata_uni <- rbind(metadata_uni,uni_sub) #store to dataframe
}
```

#Data manipulation:
```{r}
#create copies of genus and metabolite data:
genera_alt <- genera
mtb_alt <- mtb
#make row names sample names and remove that column:
rownames(genera_alt) <- genera_alt[,1 ]
genera_alt <- genera_alt[,-1]
rownames(mtb_alt) <- mtb_alt[,1 ]
mtb_alt <- mtb_alt[,-1]

#select only data on the baseline samples:
uni_select <- as.vector(metadata_uni$Sample)
genera_alt <- genera_alt[uni_select,]
mtb_alt <- mtb_alt[uni_select,]

#make all data into dataframes and input numeric:
genera_alt <- as.data.frame(genera_alt)
genera_alt[] <- lapply(genera_alt, as.numeric)
mtb_alt <- as.data.frame(mtb_alt)
mtb_alt[] <- lapply(mtb_alt, as.numeric)

#make the column names the genus names:
colnames(genera_alt) <- genus_names$`Genus names`

#create new column 'Disease Group' which is just IBD or Control:
for (val in 1:51){
  if (metadata_uni$Study.Group[val] == "nonIBD"){
    metadata_uni$Disease.Group[val] <-  "Control"
  }
  else {
    metadata_uni$Disease.Group[val] <- "IBD"
  }
}
```

#Extract the 74 selected metabolites from the main dataset:
```{r}
diff_gen <- as.vector(genus_names[1:74, 3]) #names stored in the third column of genus name dataset
diff_gen <- unlist(diff_gen)
genera_diff <- genera_alt[c(diff_gen)]
```

#Extract the 14 metabolites from the main dataset found in this one:
```{r}
#extract name of smaple of metabolites:
index_mtb <- c(15,17,38,72,127,169,173,221,264,274,280,290,308,493)
select_mtb <- mtb.map[c(index_mtb),]
com_sel <- select_mtb$Compound

#extract those compounds from the numerical dataset:
mtb_imp <- mtb_alt[,c(com_sel)]
```

#Extract data:
```{r}
#Genera:
write.csv(genera_diff, "hmp_genera.csv", row.names = TRUE)
write.csv(metadata_uni, "hmp_meta.csv", row.names = TRUE)
write.csv(mtb_imp, "hmp_metabolites.csv", row.names = TRUE)
```

#Analysing Baseline Characteristics:
```{r}
#Compare categorical variables: 
CreateCatTable(vars = c("Gender", "Antibiotics", "race"), strata = c("Disease.Group"), data = metadata_uni, test = T, addOverall = T)

#Compare age across the disease groups:
tapply(metadata_uni$consent_age, metadata_uni$Disease.Group,summary,na.rm=TRUE)
mean(metadata_uni$consent_age, na.rm=TRUE)
sd(metadata_uni$consent_age, na.rm=TRUE)
median(metadata_uni$consent_age, na.rm=TRUE)
quantile(metadata_uni$consent_age, prob = c(0.25, 0.5, 0.75), na.rm = TRUE)
t_test <- t.test(consent_age ~ Disease.Group,
           data = metadata_uni)
(t_test$p.value)
```

#Statistical Analysis:
Compare the abundance of each selected genus between the two groups with a 
Mann-Whitney Wilcoxon:
```{r}
ibd_index <- which(metadata_uni$Disease.Group == "IBD")
control_index <- which(metadata_uni$Disease.Group == "Control")
man_p_gen <- c(rep(0,74))
for (feat in 1:74){
  x <- genera_diff[c(ibd_index), feat]
  y <- genera_diff[c(control_index), feat]
  man_wit <- wilcox.test(x, y)
  man_p_gen[feat] <- man_wit$p.value
}
man_p_gen_adj <- p.adjust(man_p_gen, method = 'hochberg', n = length(man_p_gen))
man_p_gen_adj
```

Compare the abundance of each selected metabolite between the two groups with a 
Mann-Whitney Wilcoxon:
```{r}
ibd_index <- which(metadata_uni$Disease.Group == "IBD")
control_index <- which(metadata_uni$Disease.Group == "Control")
man_p_mtb <- c(rep(0,14))
for (feat in 1:14){
  x <- mtb_imp[c(ibd_index), feat]
  y <- mtb_imp[c(control_index), feat]
  man_wit <- wilcox.test(x, y)
  man_p_mtb[feat] <- man_wit$p.value
}
man_p_mtb_adj <- p.adjust(man_p_mtb, method = 'hochberg', n = length(man_p_mtb))
man_p_mtb_adj
```

Identify any genus differentially abundant between disease groups in this dataset:
```{r}
#remove all genus with greater than 80% sparsity:
sparsity <- c(rep(0, 9694))
for (i in 1:ncol(genera_alt)) {
    sparsity[i] <- (sum(genera_alt[, i] == 0)/51)*100
}
remove <- which(sparsity > 80)
genera_alt <- genera_alt[-c(remove)]

#complete Mann-Whitney Wilcoxon between the two groups:
ibd_index <- which(metadata_uni$Disease.Group == "IBD")
control_index <- which(metadata_uni$Disease.Group == "Control")
man_p <- c(rep(0,2072))
for (feat in 1:2072){
  x <- genera_alt[c(ibd_index), feat]
  y <- genera_alt[c(control_index), feat]
  man_wit <- wilcox.test(x, y)
  man_p[feat] <- man_wit$p.value
}
man_p_adj <- p.adjust(man_p, method = 'hochberg', n = length(man_p))
sig_feat <- which(man_p_adj < 0.05)
man_p
```
No significant genus identified. 